---
title: "Data Visualization in R with ggplot2"
author: "Abilash Cherian"
date: "5/5/2020"
output:
  html_document
---

```{r setup,include=FALSE}
knitr::opts_chunk$set(error =TRUE)
```

# Why use R and ggplot2 #

- Reproducibility

- Create many variations and iterations without having to start from scatch

**There are several plotting systems available in R**

- Base graphics

- lattice

- plolty

- ggplot2

Many users happily use base graphics and lattice, and it is possible to create excellent visualizations.

Personally, I find ggplot2 more intuitive. There are more libraries that helps make specialized plots in R

# Resources to learn #

+ *ggplot2:Elegant graphics for Data Analysis* by Hadley Wickham
+ *R Graphics Cookbook* by Winston Chang
+ Online documentaion: https://www.rdocumentation.org/packages/ggplot2/versions/3.3.0


```{r eval=FALSE}
# Let's start with Installing the below packages
install.packages(c('ggplot','dplyr','gapminder'))
```


```{r eval=FALSE}
#  Load the packages
library(ggplot2)
library(dplyr)
library(gapminder)
```

# Understanding ggplot2's Grammer of Graphics

+ Articulated by Leland Wilkinson: fundamentals features that make up statistical graphics.

+ Hadley Wicham: build the layered grammar of graphics in ggplot2

###  Most fundamental component of Data Visualization is Data

```{r}
head(mtcars)

```

+ **ggplot2 expects data to be stored in a data frame with coloums as variables**
+ **Next is Mapping: Aesthestics**  

      - X = weight of the Vehicle  
      - Y = Miles per Gallon  
      
+ **Geometrics Objects ('Geoms')**

      - Visual elements you see on the plot  
      - Points, lines, shapes etc.  

```{r echo=FALSE,fig.height=3}
library(ggplot2)
ggplot(mtcars,aes(wt,mpg))+
  geom_point()
```

      
+ **Statistical Transformations ('STATS')**  

      - Transforms data, usually to summarize it
      - Fit data to smoother
      
```{r, echo=FALSE,fig.height=3,warning=FALSE}
library(ggplot2)
ggplot(mtcars,aes(wt,mpg))+
  geom_point()+
  stat_smooth(method = 'loess',
              formula = 'y ~ x')
```

+ **Facets**

    - Breaks up data into subsets to visualize as small multiples   

+ **Scales**

    - Controls how data values are mapped to aesthestics
    - Color Scale
    - Long transform axis
    

+ **Coordinate Systems ("COORD')**

    - controls how data coordinates are mapped to the visual plane of the graphic
    
### Summarize gglot2's grammer of graphics

1. Data
2. Aesthetics 
3. Geoms
4. Stats
5. Facets
6. Scales
7. Coordinates

# Data Manipulation

#### Using Pipe Operator makes easy data manipulation **%>%**

```{r, eval=TRUE}
# Load the packages
library(gapminder)
library(dplyr)

```


```{r}
#  summarizing the median population

gapminder %>% 
  filter(year == 2007) %>% 
  group_by(continent) %>% 
  summarise(medpop = median(pop)) %>% 
  head()
```

### Let's plot the graphics

```{r}

# First filter the data

gap_07 <- gapminder %>% 
  filter(year == 2007)

#  lets check the first 10 rows of the data

head(gap_07)
```

### Plotting using geom_point()

```{r}
ggplot(gap_07,aes(x = gdpPercap, y = lifeExp))+
  geom_point()
```

### Plotting using geom_bar()

```{r}
ggplot(gap_07,aes(x = continent))+
  geom_bar()
```

```{r}
# Plotting the median of the population

gap_med <- gapminder %>% 
  filter(year == 2007) %>% 
  group_by(continent) %>% 
  summarise(medpop = median(pop))

ggplot(gap_med,aes(x = continent, y = medpop))+
  geom_bar(stat = 'identity')
```


#### Let's Save the plot using ggsave()

```{r,error=FALSE}
# Create an object 'p'

p <- ggplot(gap_07,aes(x = continent,y = gdpPercap))

p + geom_line() +
  geom_point() +
  ggsave('ggplot2.image.png',height = 6,width = 8)

```

# Viewing Data Distributions

```{r}
# Applying geom_histogram()
ggplot(gap_07,aes(x = gdpPercap))+
  geom_histogram(bins = 15)
```

```{r}
# Using geom_density()
ggplot(gap_07,aes(x = gdpPercap))+
  geom_density()
```
```{r}
# using boxplot and jitter point
ggplot(gap_07,aes(x = continent, y = lifeExp))+
  geom_boxplot()+
  geom_jitter(width = 0.5,alpha = 0.5)
```

# Additional visual elements: Color, Shape & Size 
 
```{r}

# Adding color and size

ggplot(gap_07, aes(x = gdpPercap, y = lifeExp,
                   color = continent, size = pop))+
  geom_point()

```

```{r}
gap_pop <- gapminder %>% 
  group_by(continent, year) %>% 
  summarise(pop = mean(pop))

ggplot(gap_pop,aes(x = year, y = pop, color = continent))+
  geom_line()
```

## Adding Facets (Small Multiples)
 
```{r}
gap_life <- gapminder %>% 
  group_by(continent, year) %>% 
  summarise(lifeExp = mean(lifeExp))

# adding facet_grid

ggplot(gap_life,aes(x = year, y = lifeExp))+
  geom_line()+
  facet_grid(continent ~ .)

#  Adding facet_wrap

ggplot(gap_life,aes(x = year, y = lifeExp))+
  geom_line()+
  facet_wrap(~ continent)
```

## Smoothing Functions

```{r,warning = FALSE}
ggplot(gap_07, aes(x = gdpPercap, y = lifeExp))+
  geom_point()+
  geom_smooth(method = 'loess')

# Let's add some more functionalities

ggplot(gap_07, aes(x = gdpPercap, y = lifeExp))+
  geom_point()+
  facet_wrap(~ continent) +
  geom_smooth(aes(color = continent))



```


# Customizing your Plots

## Axis Scales

```{r}
ggplot(gap_07, aes(x = gdpPercap, y = lifeExp))+
  geom_point()+
  scale_x_continuous(trans = 'log10')

# Or you can use scale_x_log10() function

ggplot(gap_07, aes(x = gdpPercap, y = lifeExp))+
  geom_point()+
  scale_x_log10()+
  # Setting Y axis limits
  scale_y_continuous(limits = c(0,95))

```

## Changing Color Scales

```{r}
ggplot(gap_07,aes(x = gdpPercap, y = lifeExp,
                  color = continent))+
  geom_point()+
  scale_x_log10()+
  scale_color_brewer(palette = 'Dark2')

#  changing color using geom_point()

ggplot(gap_07,aes(x = gdpPercap, y = lifeExp))+
  geom_point(color = 'red')

```

## Themes 

**Theme Functions**

- Elements
- Elements functions
- theme()
- Complete themes

```{r}
ggplot(gap_07,aes(x = gdpPercap, y = lifeExp,
                  color = continent))+
  geom_point()+
  scale_x_log10()+
  theme_bw()

# Using dark

ggplot(gap_07,aes(x = gdpPercap, y = lifeExp,
                  color = continent))+
  geom_point()+
  scale_x_log10()+
  theme_dark()

#  Using Void function

ggplot(gap_07,aes(x = gdpPercap, y = lifeExp,
                  color = continent))+
  geom_point()+
  scale_x_log10()+
  theme_void()
```


You can also use `ggthemes` to get amazing themes for your project

```{r}
install.packages('ggthemes')

# Load the library(ggthemes)

library(ggthemes)
```

```{r}
ggplot(gap_07,aes(x = gdpPercap, y = lifeExp,
                  color = continent))+
  geom_point()+
  scale_x_log10()+
  theme_tufte()

# Using Solarized theme

ggplot(gap_07,aes(x = gdpPercap, y = lifeExp,
                  color = continent))+
  geom_point()+
  scale_x_log10()+
  theme_solarized()+
  scale_colour_solarized(accent = 'blue')

# Finally you can use base R theme functions

ggplot(gap_07,aes(x = gdpPercap, y = lifeExp,
                  color = continent))+
  geom_point()+
  theme_base()

```

# Customizing Titles, Axis, Labels and Legends  


Know more at https://ggplot2.tidyverse.org/reference/ggtheme.html

```{r}
ggplot(gap_07,aes(x = gdpPercap, y = lifeExp,
                  color = continent))+
  geom_point(size = 2)+
  theme_light()+
  theme(legend.position = 'bottom',
        legend.key = element_blank(),
        axis.text = element_text(size = 12))

# You can give No Legend

ggplot(gap_07,aes(x = gdpPercap, y = lifeExp,
                  color = continent))+
  geom_point(size = 2)+
  theme_light()+
  theme(legend.position = 'none')

# Adding New Labels
ggplot(gap_07,aes(x = gdpPercap, y = lifeExp,
                  color = continent))+
  geom_point(size = 2)+
  theme_light()+
  labs(x = 'Per Capita GDP',
       y = 'Life Expectancy',
       title = '2007 Life Expectancy and GDP',
       color = 'Continent')


```



